### 2020-04-14 manuscript/manuscript.rmd
pass at methods

### Sensitive test of CAT categories {place holder for sensitive results}
The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices. CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix [@Lartillot:2004dq] . Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Table XX). The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx). Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results, potentially indicating that the C60 models might be already overfitting to most matrices. These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].

@@

### Bias of data-recoding
There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies. They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices. Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig. 3B). Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi, we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 5). Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).

@@

All tree files, intermidiate results and bioinformatic commands associated with this study are available at https://github.com/caseywdunn/animal_root.

### Data selection and wrangling (Ben)

We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository. We manually edited some minor formatting to make the batch processing of the matrices *en masse*, e.g. standardizing the formatting of charset blocks. All changes made are tracked with git.

### Matrix comparison and annotation (Ben)

#### Taxon name reconciliation(Ben)

We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix. We also use a table where manual entries were needed (Supplementary Table 5, reconciliation/taxonomy_info/manual_taxonomy_map.tsv), e.g. authors of original matrix indicate species name in original manuscript. For a table summarizing all samples and their new or lengthened names, see Supplementary Table 6 (reconciliation/taxonomy_info/taxon_table.tsv).

#### Sequence Comparisons (Ben)

Using the partition files for each matrix, we isolated each sequence for each taxon from each partition. Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.

We used DIAMOND [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters. We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 78.0, eValue < 1e-15, no self->self). We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set. We also ran a BLAST+ blastp search against the SwissProt [@boeckmann2003swiss] database, filtering such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).

#### Partition Network (Ben)

@@

We constructed a network with Python and NetworkX [@Hagberg:2008] v2.2 where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions. We extracted each connected component from this network. We further split these components if the the most connected node (i.e. most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components. We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes. See Supplementary Table 7 [partition_network_summary table] for a summary tally of each part of the comparison.

@@

#### Phylogenetic analyses in IQtree
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g. site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree v1.6.7 [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing. For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like Hejnol2009, Borrowiec2015 and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

We then reanalyzed each matrix under a panel of evolutionary models, including WAG+G, GTR+G, poisson+C60+G and associated best-fit model identified above. Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis. Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C60 model or CAT site-heterogeneous models. To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models. For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*. 2009 [@hejnol2009assessing], since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept). It should be noted that the main conclusion of Simion *et al*. was also based on selection of 25% of genes for their jackknife approach.

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C60 models in IQtree and Poisson-CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree. Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes MPI v1.8 [@Lartillot:2013fg]. To minimize computational burden, CAT-GTR models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.

For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above. For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. For those matrices that were not converged, PhyloByes analyses were run for at least two weeks. We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.

#### phylogenetic distribution of support
To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices. We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta). For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG and LG models. The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans. The constraint trees were conducted in a customized R script. The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al. 2017 [@shen2017contentious].

For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [Place holder for Xingxing]

#### Sensitive analyses with different number of substitutional categories
To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60). To compare the results between Poisson-CAT and GTR-CAT and minimize computational burden, GTR-CAT and GTR-CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full. All phyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers fo categories were used.

@@

Each new matrix was analyzed with PhyloBayes-mpi. Analyses were run for 1000 generations, and a 200 generation burnin applied. The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.

@@

### Matrix mapping (Ben)

Taxa and partition correspondence across manuscripts was assessed by comparing all sequences for each taxon in each partition across all matrices with diamond blast. Based on inspection of sequence similarity, we excluded all comparisons with less than 99% identity and greater than 10^-25^ e-value.

### Taxa comparison across matrices (Ben)

The primary intent of comparing taxa across matrices was to validate our taxon name reconciliation across studies.

We first considered pairwise similarity between the same species from different matrices in different studies.

@@

We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam cluster.

@@

- CAT site heterogeneous [@Lartillot:2004dq]. Each site is assigned to a class with its own equilibrium frequencies. The number of classes, assignment of sites to classes, and equilibrium frequencies within the data are all estimated in a Bayesian framework.

@@

The data matrix was constructed using a semi-automated approach. Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplications. Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors). The remaining gene trees with more than one sequence for any taxon were then manually inspected. If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded. If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded. Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix. Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships. Unstable taxa were identified with leaf stability indices [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix. This resulted in the 64-taxon matrix that is the focus of most of their analyses. Phylogenetic analyses were conducted under the Poisson+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].

@@

Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals by added 9 new species. Th data matrix were assembled based on The phylogenetic analysis using Poisson-CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade. Gene trees were then constructed, and potentially paralogs were removed by a bootstrap threshold of 70. Ambiguously aligned regions were trimemed and only genes sampled for at least two-thirds of species were retained. The phylogenetic analyses were conducted under the Poisson+CAT model in phylobayes [@Lartillot:2004dq].

@@

All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a Bayesian framework and with bootstrapping. All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors. Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS). With broader Opisthokonta sampling, support for Porifera-sister is 84% PP. This is an interesting case where increased outgroup sampling leads to increased support for Porifera-sister.

@@

Orthologs were constructed using the bioinfomatics pipeline OrthoSelect [@schreiber2009orthoselect]. They also evaluated level of saturation, leaf stability of sampled taxa, compositional heterogeinity and model comparison of each matrix. By modifying gene sampling, ingroup and outgroup sampling, three major topologies related to the position of animal-root were constructed (including Porifera+Placozoa sister, Ctenophora-sister and Porifera-sister). Phylogenetic analyses were conducted under the Poisson+CAT, GTR+CAT and GTR models in phylobayes [@Lartillot:2004dq].

@@

Pisani *et al.* [@pisani2015genomic] reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013_est, Moroz2014_3d and Whelan2015 datasets. It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of the position of the root of animal phylogeny. The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction. Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in phylobayes. They found GTR-CAT and Poisson-CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice. Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.

@@

Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally. This is important given the growing interest in recoding [@Feuda:2017ew]. Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies. They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding. An alternative hypothesis is that recoding impacts phylogenetic analyses because it discards so much information. These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange. Empirical and random recodings both discard the same amount of information, but only empirical recoding reduce the impact of amino-acid frequency as intended. Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate compositional heterogeneity. Similar results would suggest that the impact of recoding is due to discarding information. Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data. we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 3B).

@@

### Partition comparison across matrices (XXBen, do we need this?)

```{r partition_mapping}

# Investigate connected components of genes (ie partitions) that include multiple members fromt he same matrix
#clusterstats_within_manuscripts = partition_map_global %>% group_by( matrix, component_number ) %>%  summarise(n=n()) %>% arrange(desc(n))
#clusterstats = partition_map_global %>% group_by( component_number ) %>% summarise(n=n()) %>% arrange(desc(n))

#clusterstats_within_manuscripts %>% print(n=100)
#clusterstats %>% print(n=100)
#partition_network_summary %>% print()

The count for a partition pair can be much larger than the number of genes in the matrix, which suggests that the count is the number of hsps rather than the number of sequences with hits.

There are `r length(unique(partition_network_summary$matrix))` matrices. A gene that is perfectly sampled would form a cluster with this size. To our surprise, very few clusters, though, are this size. This suggests that intersection of genes between matrices is extreme low. This finding suggested genes used for exploring the position of the root of the animal phylogeny

>>>>

### Sensitive test of CAT categories {place holder for sensitive results}
The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices. CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix. Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Table XX). The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx). Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results, potentially indicating that the C60 models might be already overfitting to most matrices. These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].

@@

### Bias of data-recoding
There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies. They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices. Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig. 3B). Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 5). Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).

@@

All tree files, intermidiate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publicaton.

### Data selection and wrangling

We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository. We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks. All changes made are tracked with git.

### Matrix comparison and annotation

#### Taxon name reconciliation

We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix. We also used a table where manual entries were needed (Supplementary Table 5, manual_taxonomy_map.tsv), e.g. authors of the original matrix indicate species name in original manuscript. For a table summarizing all samples and their new or lengthened names, see Supplementary Table 6 (taxon_table.tsv).

#### Sequence comparisons

Using the original partition files for each matrix, we separated each sequence for each taxon from each partition. Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.

We used DIAMOND v0.9.26 [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters. We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 50.0, eValue < 1e-5, no self vs self). We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set. We also ran a BLAST+ v2.8.1 [@Camacho:2009] blastp search against the SwissProt [@boeckmann:2003swiss] database, filtering results such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).

#### Partition network

@@

We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions. We extracted each connected component from this network. We further split these components if the the most connected node (i.e. most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components. We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes. See Supplementary Table 7 [XX partition_network_summary table] for a summary tally of each part of the comparison.

@@

#### Phylogenetic analyses in IQtree
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g. site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQtree v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing. For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

We then reanalyzed each matrix under a panel of evolutionary models, including WAG+G, GTR+G, poisson+C60+G and associated best-fit model identified above. Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis. Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C60 model or CAT site-heterogeneous models. To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models. For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*. 2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept). It should be noted that the main conclusion of Simion *et al*. was also based on selection of 25% of genes for their jackknife approach.

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C60 models in IQtree and Poisson-CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree. Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes-MPI v1.8. To minimize computational burden, CAT-GTR models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.

For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above. For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks. We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.

#### Phylogenetic distribution of support
To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices. We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta). For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG and LG models. The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans. The constraint trees were conducted in a customized R script. The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al. 2017 [@shen2017contentious].

For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]

#### Sensitive analyses with different number of substitutional categories
To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60). To compare the results between Poisson-CAT and GTR-CAT and minimize computational burden, GTR-CAT and GTR-CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full. All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers fo categories were used.

@@

Each new matrix was analyzed with PhyloBayes-MPI. Analyses were run for 1000 generations, and a 200 generation burnin applied. The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.

@@

@@

We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam HPC cluster.

@@

- CAT site heterogeneous. Each site is assigned to a class with its own equilibrium frequencies. The number of classes, assignment of sites to classes, and equilibrium frequencies within the data are all estimated in a Bayesian framework.

@@

The data matrix was constructed using a semi-automated approach. Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplications. Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors). The remaining gene trees with more than one sequence for any taxon were then manually inspected. If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded. If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded. Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix. Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships. Unstable taxa were identified with leaf stability indices [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix. This resulted in the 64-taxon matrix that is the focus of most of their analyses. Phylogenetic analyses were conducted under the Poisson+CAT model in PhyloBayes, and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].

@@

Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals by added 9 new species. Th data matrix were assembled based on The phylogenetic analysis using Poisson-CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade. Gene trees were then constructed, and potentially paralogs were removed by a bootstrap threshold of 70. Ambiguously aligned regions were trimemed and only genes sampled for at least two-thirds of species were retained. The phylogenetic analyses were conducted under the Poisson+CAT model in PhyloBayes.

@@

All their analyses were conducted under the F81+CAT+Gamma model in PhyloBayes, in both a Bayesian framework and with bootstrapping. All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors. Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS). With broader Opisthokonta sampling, support for Porifera-sister is 84% PP. This is an interesting case where increased outgroup sampling leads to increased support for Porifera-sister.

@@

Orthologs were constructed using the bioinfomatics pipeline OrthoSelect [@schreiber2009orthoselect]. They also evaluated level of saturation, leaf stability of sampled taxa, compositional heterogeinity and model comparison of each matrix. By modifying gene sampling, ingroup and outgroup sampling, three major topologies related to the position of animal-root were constructed (including Porifera+Placozoa sister, Ctenophora-sister and Porifera-sister). Phylogenetic analyses were conducted under the Poisson+CAT, GTR+CAT and GTR models in PhyloBayes.

@@

Pisani *et al.* [@pisani2015genomic] reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013_est, Moroz2014_3d and Whelan2015 datasets. It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of the position of the root of animal phylogeny. The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction. Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in PhyloBayes. They found GTR-CAT and Poisson-CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice. Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.

@@

Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally. This is important given the growing interest in recoding [@Feuda:2017ew]. Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies. They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding. An alternative hypothesis is that recoding impacts phylogenetic analyses because it discards so much information. These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange. Empirical and random recodings both discard the same amount of information, but only empirical recoding reduce the impact of amino-acid frequency as intended. Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate compositional heterogeneity. Similar results would suggest that the impact of recoding is due to discarding information. Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data. we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 3B).

@@
